Dissociation between vertebrae number and body size in Australian blindsnakes

### 2024-08-23 manuscript.Rmd
discussion paragraph 1

In another predominantly burrowing clade, caecilians, the number of vertebrae is positively correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversity in the underground environment. 
Pleomerism has been observed in typhlopidae, however, some notable exceptions indicate different mechanisms for gigantism [@headDissociation2007]. For instance, Head and Polly [-@headDissociation2007] found that some of the largest species of blindsnakes in the African continent (e.g. *Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae. Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001]. This is reflected in the relatively short tails of blindsnakes [Tiatragul unpublished; @szyndlarIllustrated2023]. Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer. Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "pleisomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.

@@

For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae, while most species have many more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) and has the most vertebrae, while *A. torresianus*, which is the longest, has relatively few vertebrae.

@@

***[Big summary]***

Results from our study revealed that pleomerism does not apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975]. Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size. We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide. This relationship between the proportion of vertebrae to body shape has been found in colubrids and viperids, where more elongated species tend to have a greater number of vertebrae than stouter ones [@lindellEvolution1994]. Variation in vertebral ratio can also be explained by both mean annual temperature and soil compactness. These results have important implications for understanding the macroevolutionary dynamics of continental-scale radiation of Australian blindsnakes.

***[Discuss why body shape might influence vertebrae ratio or vice-versa]***

@@

REGIONALISATION Need to mention, plus the heart position aspect; how that’s not been studied here, but it helps to understand regionalisation, which is an important next step in understanding the lack of pleomerism. Since the lack of pleomerism in *Hydrophis* sea snakes is due to their elongate forebodies. Tamika’s paper on colubrids also shows the various ways elongation occurs, using different regions. Hampton 2023 paper on snake regionalisation in greater detail

>>>>

<!-- In another predominantly burrowing clade, caecilians, the number of vertebrae is not always correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversify in the underground environment.  -->

Pleomerism has been observed in typhlopidae, however, some notable exceptions indicate different mechanisms for gigantism [@headDissociation2007]. For instance, Head and Polly [-@headDissociation2007] found that some of the largest species of blindsnakes in the African continent (e.g. *Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae. Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001]. This is reflected in the relatively short tails of blindsnakes [Tiatragul unpublished; @szyndlarIllustrated2023]. Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer. Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "plesiomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.

@@

For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which is the longest, has relatively few vertebrae.

@@

<!-- However, information about the relationship between body size and vertebrae number is unknown. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes? (3) Does vertebral ratio vary with soil compactness or mean annual temperature? -->

Our study investigated the presence of pleomerism in Australian blindsnakes and test whether deviations from pleomerism may be explained by body shape or ecological factors. Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975]. Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size. We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide. This relationship between the proportion of vertebrae to body shape has been found in colubrids and viperids, where more elongated species tend to have a greater number of vertebrae than stouter ones [@lindellEvolution1994]. We also found that vertebral ratio correlate positively with mean annual temperature and negatively with soil compactness. These results have important implications for understanding the macroevolutionary dynamics of continental-scale radiation of Australian blindsnakes.

### Absence of pleomerism in Australian blindsnakes

We found a dissociation between vertebrae number and body size in Australian blindsnakes. This is consistent with a previous study on blindsnakes from other continents, where exceptionally large species are shown to have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007]. Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004]. Unforutnately, without life-history data for blindsnakes, we cannot determine which of these mechanisms are responsible. 
Elongation may also be achieved in other ways. The regionalisation of vertebrae, where distinct regions of the vertebrae have different evolutionary pathways and morphological diversification [@hamptonRegionalization2023;@hamptonHeart2022], may explain the lack of pleomerism. For instance, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the numnber of smaller vertebrae in the pre-atrial region compared to larger vertebrae in the post-atrial region [@sherrattMacroevolution2022;@sherrattHeterochronic2019]. Similarly, colubrids have been also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024]. Another important aspect of regionalisation is the position of the heart, which has been shown to be a key marker for delimiting different regions of the vertebrae [@hamptonHeart2022;@hamptonRegionalization2023]. Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.

### Correlation between body shape and vertebrae ratio

@@
